---
title: "Illustrating the central limit theorem using simulation"
author: "Adam Acosta"
date: "02/11/2015"
output: pdf_document
---

### Overview

In this project, we aim to illustrate the Central Limit Theorem by simulating
samples from the exponential distribution, a highly right-skewed distribution,
to illustrate the power of the theorem, that the distribution of the sample
mean drawn from any underlying distribution, even a highly skewed distribution,
will tend toward approximately normal as the number of samples means tends to 
infinity. 

### Simulations

To perform the simulation, we set the random number generator seed to 1000, then
produce 1000 random vectors of size n = 40 from the exponential distribution, 
taking the mean of each vector and storing these as our data to analyze. 

```{r,echo=FALSE}
set.seed(1000)
means <- vector(mode="numeric",length="1000")
for (i in 1:1000) means[i] <- mean(rexp(40,0.2))
```

### Experimental versus Theoretical Moments

The first part of our analysis compares the summary statistics of our simulated
empirical distribution to their expected values. 

The theoretical mean and variance of the exponential distribution with 
lambda = 0.2 are known from formula, the mean being 1/lambda and the variance
1/lambda^2. We compare these below to the experimentally simulated mean and
variance. 

```{r, echo=FALSE,results="asis",warning=FALSE,message=FALSE}
library(xtable)
library(dplyr)
tableData <- data.frame(cbind(c(mean(means),var(means)),
                              c(1/0.2,(1/0.2^2)/sqrt(1000))))
names(tableData) <- c("Experimental","Theoretical")
tableData <- mutate(tableData,Delta=Theoretical - Experimental)
row.names(tableData) <- c("Mean","Variance")
table <- xtable(tableData)
caption(table) <- "Difference between the experimental and theoretical
                   mean and variance of 1000 simulated means of the exponential
                   distribution with parameters n = 40 and lambda = 0.2."
print(table, type="latex",comment=FALSE)
```

We see that, in each case, the simulated value is very close to the 
theoretical value. 

### Visualizing the Distribution

The simplest way in which  we can visualize the empirical distribution of our 
simulation is by using a histogram. Figure 1 shows this histogram.

```{r,echo=FALSE}
hist(means,breaks=20,freq=F,col="red",xlab="",ylab="",main="")
title(main="Relative Frequency of Sample Mean")
title(sub="Figure 1. Relative frequency of the means of one thousand
      simulated samples of size n = 40 from the exponential distribution.")
```

Another way to come directly compare our empirical distribution to the normal is 
to overlay the kernel density with the normal distribution to give a visual idea 
of how closely they match. This is shown below in Figure 2. 

```{r,echo=FALSE}
x <- seq(2,8,length=1000)
px <- dnorm(x,mean=5)
plot(density(means),main="",xlab="",ylab="",xlim=c(2,8))
polygon(density(means),col="red",border="red")
lines(x,px,col="blue")
legend("topright",legend=c("Simulated Values","Normal Values"),
       col=c("red","blue"),lty=1)
title(main="Density of Sample Mean")
title(sub="Figure 2. Probability density function of simulated sample means
      overlaid with the normal distribution.")
```

Finally, the preferred visual means of testing how close a distribution is to 
normal is the quantile-quantile plot, where we plot the theoretical quantiles (if 
the underlying distribution is actually normal) against the observed quantiles. 
We see below that the empirical distribution of the sample mean does appear to be
very close to normal. 

```{r,echo=FALSE}
qqnorm(means,col="red",pch=20,main="",xlab="",ylab="")
qqline(means,col="blue",lwd=2)
legend("topleft",legend=c("Simulated Quantiles","Normal Quantiles"),
       col=c("red","blue"),lty=1)
title(main="q-q Plot of Sample Means")
title(sub="Figure 3. Quantile-quantile plot of theoretical quantiles from the
      normal distribution against observed sample quantiles.")
```

### Discussion

What we see with our simple exploratory analysis is that our sample distribution
is somewhat right-skewed, in that the right tail is much longer than the left. We 
can see in both the density plot as well as the q-q plot that the sample diverges 
furthest from normal in the left tail, where our sample quantiles are quite a bit 
greater than the equivalent normal quantiles.

This is not surprising, as the Central Limit Theorem only specifies that the
distribution of means of any arbitrary distribution tends to approximately normal 
as n tends to infinity. The closeness of fit will always be greater, however,
when the underlying distribution we draw samples from is itself a symmetric
distribution. 

Sampling from the exponential distribution for our simulations, this is not the
case. In particular, no value drawn from an exponential distribution can
ever be negative. Thus, you will never get a negative mean, or even a mean of 
zero. On the other hand, the normal distribution contains non-negative
probabilities for every possible value, all the way to negative infinity. This
explains why the left tails are different, as the left tail of our sample
distribution cuts off at zero, whereas the left-tail of a true normal 
distribution extends forever. 

Nonetheless, we do see, both from the overlaid density plots and the q-q plot,
that the distribution can be characterized as approximately normal, certainly
far more so than the exponential distribution itself. Compared to the expoential
distribution itself, the distribution of sample means is strikingly more
bell-curve shaped. 

### Appendix A. Code for Simulation

```{r,eval=FALSE}
set.seed(1000)
means <- vector(mode="numeric",length="1000")
for (i in 1:1000) means[i] <- mean(rexp(40,0.2))
```

### Appendix B. Code for Producing Table

```{r,eval=FALSE}
library(xtable)
library(dplyr)
tableData <- data.frame(cbind(c(mean(means),var(means))),c(1/0.2,1/0.2^2)))
names(tableData) <- c("Theoretical","Experimental")
mutate(tableData,Delta=Theoretical - Experimental)
row.names(tableData) <- c("Mean","Variance")
table <- xtable(tableData)
caption(table) <- "Table 1. Difference between the experimental and theoretical
                   mean and variance of 1000 simulated means of the exponential
                   distribution with parameters n = 40 and lambda = 0.2."
print(table)
```

### Appendix C. Code for Producing Plots

```{r,eval=FALSE}
# Figure 1
hist(means,breaks=20,freq=F,col="red",xlab="",ylab="",main="")
title(main="Relative Frequency of Sample Mean")
title(sub="Figure 1. Relative frequency of the means of one thousand
      simulated samples of size n = 40 from the exponential distribution.")
```

```{r, eval=FALSE}
# Figure 2
x <- seq(2,8,length=1000)
px <- dnorm(x,mean=5)
plot(density(means),main="",xlab="",ylab="",xlim=c(2,8))
polygon(density(means),col="red",border="red")
lines(x,px,col="blue")
legend("topright",legend=c("Simulated Values","Normal Values"),
       col=c("red","blue"),lty=1)
title(main="Density of Sample Mean")
title(sub="Figure 2. Probability density function of simulated sample means
      overlaid with the normal distribution.")
```

```{r,eval=FALSE}
# Figure 3
qqnorm(means,col="red",pch=20,main="",xlab="",ylab="")
qqline(means,col="blue",lwd=2)
legend("topleft",legend=c("Simulated Quantiles","Normal Quantiles"),
       col=c("red","blue"),lty=1)
title(main="q-q Plot of Sample Means")
title(sub="Figure 3. Quantile-quantile plot of theoretical quantiles from the
      normal distribution against observed sample quantiles.")
```