---
title: "The impact of Vitamin C on tooth growth in guinea pigs"
author: "Adam Acosta"
date: "02/11/2015"
output: pdf_document
---

### Overview

The aim of this report is to determine from experimental data whether vitamin
C can promote tooth growth in guinea pigs. This will be done using one of the
built-in datasets from R and a variety of exploratory analyses and statistical
significance tests. 

### Description of Dataset

The ToothGrowth dataset from the R datasets package contains the tooth lengths
of 60 guinea pigs, 10 each at three different dose levels of Vitamin C, delivered
through two methods: orange juice and ascorbic acid. This gives us six unique
treatment groups, with no control. 

```{r,echo=FALSE}
library(datasets)
data(ToothGrowth)
data <- ToothGrowth
```

### Exploratory Analysis

We can begin our exploration of the data by calculating the sample mean and
standard deviation of each treatment group. Additionally, we can plot the 
estimated kernel density of each dosage level overlaid with each other. This
should give us some indication of the difference, if any, between the true
underlying distributions. This simple summary analysis seems to suggest a
significant treatment effect. 

```{r,echo=FALSE,results="asis",message=FALSE}
library(dplyr)
library(xtable)
means <- vector(mode="numeric",length=6)
sds <- vector(mode="numeric",length=6)
means[1] <- mean(filter(data,supp=="VC",dose==0.5)$len)
means[2] <- mean(filter(data,supp=="VC",dose==1.0)$len)
means[3] <- mean(filter(data,supp=="VC",dose==2.0)$len)
means[4] <- mean(filter(data,supp=="OJ",dose==0.5)$len)
means[5] <- mean(filter(data,supp=="OJ",dose==1.0)$len)
means[6] <- mean(filter(data,supp=="OJ",dose==2.0)$len)
sds[1] <- sd(filter(data,supp=="VC",dose==0.5)$len)
sds[2] <- sd(filter(data,supp=="VC",dose==1.0)$len)
sds[3] <- sd(filter(data,supp=="VC",dose==2.0)$len)
sds[4] <- sd(filter(data,supp=="OJ",dose==0.5)$len)
sds[5] <- sd(filter(data,supp=="OJ",dose==1.0)$len)
sds[6] <- sd(filter(data,supp=="OJ",dose==2.0)$len)
table <- data.frame(cbind(means,sds))
names(table) <- c("Mean","St Dev")
row.names(table) <- c("Ascorbic Acid 0.5","Ascorbic Acid 1.0","Ascorbic Acid 2.0"
                      ,"Orange Juice 0.5","Orange Juice 1.0","Orange Juice 2.0")
table <- xtable(table)
caption(table) <- "Mean and standard deviation (n=10) of tooth lengths of Guinea
                   pigs treated with different doses of Vitamin C, using pure
                   ascorbic acid and orange juice as delivery vehicles."
print(table,comment=F,type="latex")
```

```{r,echo=FALSE,fig.height=3.5}
par(mfrow=c(1,2))
boxplot(len~dose,filter(data,supp=="VC"),col=c("red","green","blue"))
title(main="Ascorbic Acid")
boxplot(len~dose,filter(data,supp=="OJ"),col=c("red","green","blue"))
title(main="Orange Juice")
```

### Quantitative Analysis

Continuing on, we can quantitively evaluate the difference between treatment
groups and estimate the magnitude and significance of this difference. 

The first thing we do is construct 95% confidence intervals, using the student's
t-distribution, for the difference between the mean tooth length in guinea pigs 
treated with different doses of vitamin C. We assume the values are not paired
and that the underlying population variance is equal. We did see above that the
sample variance is not equal for each treatment group, but we will here assume
that this is an effect of the small sample sizes. 

In each case, the entire interval lies well below zero, which suggests there is 
indeed a treatment effect. The p-value in each case is so small as to round to 
zero. 

```{r,echo=FALSE,results="asis"}
half_to_one <- unique(t.test(len~dose,data,subset=c(1:20,31:50),
                             var.equal=TRUE)$conf.int)
half_to_one[3] <- t.test(len~dose,data,subset=c(1:20,31:50),
                         var.equal=TRUE)$p.value
one_to_two <- unique(t.test(len~dose,data,subset=c(11:30,41:60),
                            var.equal=TRUE)$conf.int)
one_to_two[3] <- t.test(len~dose,data,subset=c(11:30,41:60),
                        var.equal=TRUE)$p.value
half_to_two <- unique(t.test(len~dose,data,subset=c(1:10,21:40,51:60),
                             var.equal=TRUE)$conf.int)
half_to_two[3] <- t.test(len~dose,data,subset=c(1:10,21:40,51:60),
                         var.equal=TRUE)$p.value
ints <- data.frame(rbind(half_to_one,one_to_two,half_to_two))
names(ints) <- c("Lower Bound","Upper Bound","p-value")
row.names(ints) <- c("0.5 to 1.0","1.0 to 2.0","0.5 to 2.0")
newTable <- xtable(ints)
caption(newTable) <- "95\\% confidence intervals for the difference of mean tooth
                      length when changing dosage."
print(newTable,comment=F,type="latex")
```

Additionally, we can perform the same procedure to construct a 95% confidence
interval for the difference in mean tooth length depending upon the method of
delivery. We see in this case that the interval includes 0, with a p-value 
greater than 0.5, suggesting the data here is inconclusive.

```{r,echo=FALSE,results="asis"}
only <- data.frame(rbind(unique(t.test(len~supp,data,
                                       var.equal=TRUE)$conf.int)))
only[,3] <- t.test(len~supp,data)$p.value
names(only) <- c("Lower Bound","Upper Bound","p-value")
row.names(only) <- c("")
suppTable <- xtable(only)
caption(suppTable) <- "95\\% confidence interval for the difference of mean tooth
                      length when changing delivery method."
print(suppTable,comment=F,type="latex")
```

```{r,echo=FALSE,results="asis"}

```

### Discussion

The data, taken at face value, suggest a treatment effect significant at p < 0.05
of Vitamin C on tooth growth in guinea pigs, but no effect from varying the 
delivery method. We may additionally observe that the treatment appears to
possibly taper with increasing dosage; that is, the 95% confidence interval
for the difference between 0.5 and 1.0 is largely greater in magnitude than
the 95% confidence interval for the difference between 1.0 and 2.0. 

However, we must note that we do not know either the sampling
or experimental methodology and how the researchers mitigated against confounding
variables. Additionally, we are working with very small samples here, as
demonstrated by the great variety in shape of the kernel densities. Finally,
there is no control group. 

Nonetheless, this report serves as an illustration of how to use basic
exploratory and inferential techniques to draw inferences about population
characteristics from sampling and experimentation. Additionally, using the
student's t-distribution allows us to draw inferences from small samples,
provided we can be reasonably assured our data are drawn from a symmetric
population distribution. 

In this case, our box plots do not necessarily suggest symmetric distributions,
but we are nonetheless making that assumption, simply given what we know about
tooth growth, that we would expect lengths to be normally distributed. Finally,
to accept these conclusions as valid, we must also note that the ascorbic acid
boxplot did indicate a single outlier that we did not exclude from the analysis,
which we justify by the small sample size, assuming these samples are not 
sufficiently large to rule out any values as outliers. 

### Appendix A. Code for Tables

```{r,eval=FALSE}
# mean and standard deviation of treatment groups conditioned on dosage
library(dplyr)
library(xtable)
means <- vector(mode="numeric",length=6)
sds <- vector(mode="numeric",length=6)
means[1] <- mean(filter(data,supp=="VC",dose==0.5)$len)
means[2] <- mean(filter(data,supp=="VC",dose==1.0)$len)
means[3] <- mean(filter(data,supp=="VC",dose==2.0)$len)
means[4] <- mean(filter(data,supp=="OJ",dose==0.5)$len)
means[5] <- mean(filter(data,supp=="OJ",dose==1.0)$len)
means[6] <- mean(filter(data,supp=="OJ",dose==2.0)$len)
sds[1] <- sd(filter(data,supp=="VC",dose==0.5)$len)
sds[2] <- sd(filter(data,supp=="VC",dose==1.0)$len)
sds[3] <- sd(filter(data,supp=="VC",dose==2.0)$len)
sds[4] <- sd(filter(data,supp=="OJ",dose==0.5)$len)
sds[5] <- sd(filter(data,supp=="OJ",dose==1.0)$len)
sds[6] <- sd(filter(data,supp=="OJ",dose==2.0)$len)
table <- data.frame(cbind(means,sds))
names(table) <- c("Mean","St Dev")
row.names(table) <- c("Ascorbic Acid 0.5","Ascorbic Acid 1.0","Ascorbic Acid 2.0"
                      ,"Orange Juice 0.5","Orange Juice 1.0","Orange Juice 2.0")
table <- xtable(table)
caption(table) <- "Mean and standard deviation (n=10) of tooth lengths of Guinea
                   pigs treated with different doses of Vitamin C, using pure
                   ascorbic acid and orange juice as delivery vehicles."
print(table,comment=F,type="latex")

# confidence intervals for difference of mean tooth length conditioned on
# dosage administered
half_to_one <- unique(t.test(len~dose,data,subset=c(1:20,31:50))$conf.int)
half_to_one[3] <- t.test(len~dose,data,subset=c(1:20,31:50))$p.value
one_to_two <- unique(t.test(len~dose,data,subset=c(11:30,41:60))$conf.int)
one_to_two[3] <- t.test(len~dose,data,subset=c(11:30,41:60))$p.value
half_to_two <- unique(t.test(len~dose,data,subset=c(1:10,21:40,51:60))$conf.int)
half_to_two[3] <- t.test(len~dose,data,subset=c(1:10,21:40,51:60))$p.value
ints <- data.frame(rbind(half_to_one,one_to_two,half_to_two))
names(ints) <- c("Lower Bound","Upper Bound","p-value")
row.names(ints) <- c("0.5 to 1.0","1.0 to 2.0","0.5 to 2.0")
newTable <- xtable(ints)
caption(newTable) <- "95\\% confidence intervals for the difference of mean tooth
                      length when changing dosage."
print(newTable,comment=F,type="latex")

# confidence interval for difference of mean tooth length conditioned on
# delivery method
only <- data.frame(rbind(unique(t.test(len~supp,data)$conf.int)))
only[,3] <- t.test(len~supp,data)$p.value
names(only) <- c("Lower Bound","Upper Bound","p-value")
row.names(only) <- c("")
suppTable <- xtable(only)
caption(suppTable) <- "95\\% confidence interval for the difference of mean tooth
                      length when changing delivery method."
print(suppTable,comment=F,type="latex")
```

### Appendix B. Code for Plots

```{r,eval=FALSE}
par(mfrow=c(1,2))
boxplot(len~dose,filter(data,supp=="VC"),col=c("red","green","blue"))
title(main="Ascorbic Acid")
boxplot(len~dose,filter(data,supp=="OJ"),col=c("red","green","blue"))
title(main="Orange Juice")
```