---
title: Human casualties and economic loss resulting from severe weather events, 
  1950 to 2011
author: Adam Acosta
date: March 2015
output: html_document
---

## Synopsis

This project aims to find the severe Weather events associated with the greatest 
number of human deaths and injuries, as well as the greatest magnitude of 
economic loss, in the United States over the period from 1950 to 2011. The data 
come from the National Oceanic and Atmospheric Administration' storm database, 
which gathers together data stored in both paper and electronic format from local 
sources. 

Documentation of the data can be found at the following urls:

National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfron
t.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

This analysis indicates that tornadoes pose the greatest risk to humans, causing 
both the greatest number of deaths and the greatest number of injuries. Floods 
cause the greatest magnitude of economic loss. 

## Data Processing

### Loading Packages

This analysis will utilize functions from the R car, data.table, and ggplot2 
packages. 

```{r, packages, message = FALSE}
library(car)
library(data.table)
library(dplyr)
library(lubridate)
library(ggvis)
```

Data is hosted in bzip2 format on cloudfront. 

### Loading Data

Here, we load the relevant data. Since all we are ultimately trying to do is find 
the total human and economic damage by event type, we read in only the 7 columns 
necessary to do this, thereby saving memory space and hopefully speeding up the 
reading in of a very large file. Additionally, as there is no data for most events
except tornados prior to 1990, I filter out all events from prior to avoid 
biasing the results. 

```{r, download, cache = TRUE}
dFile <- 'http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
destfile <- 'data.csv.bz2'
if (!file.exists(destfile)) {
    download.file(dFile, destfile = destfile)
}
# Note: fread() error encountered: 
# Error in fread() : 
#  mmap'd region has EOF at the end
# read.table used instead
data <- data.table(read.table('data.csv.bz2', 
                              header = TRUE, 
                              sep = ',', 
                              nrows = 902298,
                              na.strings = '',
                              colClasses = c('NULL', 
                                             'character',
                                             rep('NULL', 5),
                                             'factor',
                                             rep('NULL', 14),
                                             rep('numeric', 3),
                                             'character', 
                                             'numeric', 
                                             'character',
                                             rep('NULL', 9)))) %>%
    mutate(DATE = as.Date(BGN_DATE, format = '%m/%d/%Y %H:%M:%S')) %>%
    filter(year(DATE) > 1989)
```

### Cleaning Data

Many of the event type observations are summaries of particular dates, rather 
than actual events. Those are discarded. Additionally, all other codes in the 
"EVTYPE" column are removed if they are obviously not weather events. Finally, 
all event types with 0 damage, 0 fatalities, and 0 injuries are removed. 

After removing these, it is necessary to turn multiple codes of single events 
into a single code. For instance, the original factor variables includes 12 
different ways of naming a water spout. We change all of these to "WATERSPOUT."

Note that this introduces a few difficulties. For instance, rain and wind, or 
snow and wind, are each distinct event types that often occur together. What I 
have done is recode anything containing rain or snow first to rain or snow, then 
recoding anything else containing wind to wind, with the rationale that it is 
more likely to be the rain or snow causing damage than the wind when each occur 
together (hurricanes are, of course, coded as hurricanes, to make the obvious 
exception).

Events including both ice and snow are recoded to ice. This could certainly be 
done either way. 

Hypothermia and hyperthermia are recoded to cold and heat, respectively, as each 
is a human injury type, not a weather event. 

Landslides and mudslides are both coded as landslides. 

```{r, clean, cache = TRUE}
data <- data %>% mutate(EVTYPE = toupper(EVTYPE)) %>%
    filter(EVTYPE != "?" & EVTYPE != "APACHE COUNTY" & EVTYPE != 
           "NO SEVERE WEATHER" & EVTYPE != "NONE" & EVTYPE != 
           "OTHER" & EVTYPE != "SEICHE" & EVTYPE != 
           "METRO STORM, MAY 26" & !grepl("ABNORMALLY",EVTYPE)) %>%
    slice(!grepl("SUMMARY", EVTYPE)) %>%
    filter(PROPDMG != 0 | CROPDMG != 0 | FATALITIES != 0 | INJURIES != 0)

data[grep("AVALANCHE", data$EVTYPE),"EVTYPE"] <- "AVALANCHE"
data[grep("BLIZZARD", data$EVTYPE),"EVTYPE"] <- "BLIZZARD"
data[grep("COLD|HYPOTHERMIA", data$EVTYPE),"EVTYPE"] <- "COLD"
data[grep("DRAUGHT|DRY|DRIEST", data$EVTYPE),"EVTYPE"] <- "DRAUGHT"
data[grep("DAM ", data$EVTYPE),"EVTYPE"] <- "DAM FAILURE"
data[grep("DOWNBURST", data$EVTYPE),"EVTYPE"] <- "DOWNBURST"
data[grep("DUST", data$EVTYPE),"EVTYPE"] <- "DUST STORM"
data[grep("FLOOD|STREAM", data$EVTYPE),"EVTYPE"] <- "FLOOD"
data[grep("FOG", data$EVTYPE),"EVTYPE"] <- "FOG"
data[grep("FREEZE", data$EVTYPE),"EVTYPE"] <- "FREEZE"
data[grep("FUNNEL", data$EVTYPE),"EVTYPE"] <- "FUNNEL"
data[grep("GLAZE", data$EVTYPE),"EVTYPE"] <- "GLAZE"
data[grep("HAIL", data$EVTYPE),"EVTYPE"] <- "HAIL"
data[grep("HEAT|HYPERTHERMIA|WARM", data$EVTYPE),"EVTYPE"] <- "HEAT"
data[grep("HURRICANE|HURRICANE|TYPHOON", data$EVTYPE),"EVTYPE"] <- "HURRICANE"
data[grep("ICE", data$EVTYPE),"EVTYPE"] <- "ICE"
data[grep("LANDSLIDE|MUD", data$EVTYPE),"EVTYPE"] <- "LANDSLIDE"
data[grep("LIGHTNING|LIGHTING|LIGNTNING", data$EVTYPE),"EVTYPE"] <- "LIGHTNING"
data[grep("MICROBURST", data$EVTYPE),"EVTYPE"] <- "MICROBURST"
data[grep("RAIN|PRECIPITATION", data$EVTYPE),"EVTYPE"] <- "RAIN"
data[grep("RIP CURRENT", data$EVTYPE),"EVTYPE"] <- "RIP CURRENT"
data[grep("SLEET", data$EVTYPE),"EVTYPE"] <- "SLEET"
data[grep("SNOW", data$EVTYPE),"EVTYPE"] <- "SNOW"
data[grep("SURGE", data$EVTYPE),"EVTYPE"] <- "STORM SURGE"
data[grep("STRONG WIND", data$EVTYPE),"EVTYPE"] <- "STRONG WIND"
data[grep("SURF", data$EVTYPE),"EVTYPE"] <- "SURF"
data[grep("THUNDERSTORM|TUNDERSTORM|TSTM|THUNDEERSTORM|THUNDERESTORM |
          THUNERSTORM|TSTMW", data$EVTYPE),"EVTYPE"] <- "THUNDERSTORM WIND"
data[grep("TORNADO|TORNDAO", data$EVTYPE),"EVTYPE"] <- "TORNADO"
data[grep("TROPICAL", data$EVTYPE),"EVTYPE"] <- "TROPICAL STORM"
data[grep("VOLCANIC", data$EVTYPE),"EVTYPE"] <- "VOLCANIC ERUPTION"
data[grep("SPOUT", data$EVTYPE),"EVTYPE"] <-"WATERSPOUT"
data[grep("WILDFIRES|WILD FIRES|WILDFIRE|WILD/FOREST", data$EVTYPE),
     "EVTYPE"] <- "WILDFIRE"
data[grep("WIND|WND", data$EVTYPE),"EVTYPE"] <- "WIND"
data[grep("WINTER STORM", data$EVTYPE),"EVTYPE"] <- "WINTER STORM"
data[grep("WINTER EVENT|WINTRY", data$EVTYPE),"EVTYPE"] <- "WINTER EVENT"
```

Property and crop damage are coded in the dataset as a mantissa exponent pair, 
with the PROPDMGEXP and CROPDMGEXP variables indicating the power of 10 to 
multiply the PROPDMG and CROPDMG variables by to get the dollar amount of loss. 
Multiple encodings are used, and we must change each of these to a numeric value 
equal to the power of 10 we need to multiply by, then carry out the 
multiplication.

Finally, I add property and crop damage into a new column to find the total loss 
associated with each event type. 

```{r, group_damage, cache = TRUE}
data$PROPDMG <- data$PROPDMG * as.numeric(recode(toupper(data$PROPDMGEXP),
                                                 "'NA'=1;'0'=1;'1'=10;
                                                 '2'=100;'3'=1000;'4'=10000;
                                                 '5'=100000;'6'=1000000;
                                                 '7'=10000000;'8'=100000000;
                                                 'H'=100;'K'=1000;'M'=1000000;
                                                 'B'=1000000000;'-'=0;'?'=0;
                                                 '+'=0; else=1", 
                                                 as.factor.result = FALSE))
data$CROPDMG <- data$CROPDMG * as.numeric(recode(toupper(data$CROPDMGEXP),
                                                 "'NA'=1;'0'=1;'2'=100;
                                                 'B'=1000000000;'K'=1000;
                                                 'M'=1000000;'?'=0;else=1", 
                                                 as.factor.result = FALSE))
data$TOTALDMG <- data$PROPDMG + data$CROPDMG
```

We then find the 10 events associated with the greatest number of injuries, the 
greatest number of deaths, and the greatest magnitude of economic loss, 
accounting for both property and crop damage. 

```{r, group_death_inj, cache = TRUE}
deaths <- aggregate(FATALITIES ~ EVTYPE, data, sum)
names(deaths) <- c("Event", "Fatalities")
deaths <- deaths[order(-deaths$Fatalities), ][1:10, ]
deaths$Event <- factor(deaths$Event, 
                       levels = deaths[order(deaths$Fatalities), "Event"])

injuries <- aggregate(INJURIES ~ EVTYPE, data, sum)
names(injuries) <- c("Event", "Injuries")
injuries <- injuries[order(-injuries$Injuries), ][1:10, ]
injuries$Event <- factor(injuries$Event, 
                         levels = injuries[order(injuries$Injuries), "Event"])

loss <- aggregate(TOTALDMG ~ EVTYPE, data, sum)
names(loss) <- c("Event", "Cost")
loss <- loss[order(-loss$Cost), ][1:10, ]
loss$Event <- factor(loss$Event, levels = loss[order(loss$Cost), "Event"])
```

## Results

### Injuries and Fatalities

We see below that tornadoes overwhelmingly cause both the most deaths and the 
most injuries. 

```{r, death_plot}
deaths %>%
    ggvis(~Event, ~Fatalities / 1e3) %>%
    layer_bars(width := 10, fill := 'steelblue') %>%
    add_axis('x', title = '', 
        properties = axis_props(
            grid = list(strokeWidth = 0), 
            ticks = list(strokeWidth = 0), 
            labels = list(angle = 30, align = 'left')
        )
    ) %>%
    add_axis('y', title = 'thousands killed', title_offset = 50)
```

```{r, injury_plot}
injuries %>%
    ggvis(~Event, ~Injuries / 1e3) %>%
    layer_bars(width := 10, fill := 'steelblue') %>%
    add_axis('x', title = '', 
        properties = axis_props(
            grid = list(strokeWidth = 0), 
            ticks = list(strokeWidth = 0), 
            labels = list(angle = 30, align = 'left')
        )
    ) %>%
    add_axis('y', title = 'thousands injured', title_offset = 50)
```

### Property Damage and Crop Loss

Additionally, flooding causes the most aggregate economic loss. 

```{r, econ_loss_plot}
loss %>%
    ggvis(~Event, ~Cost / 1e9) %>%
    layer_bars(width := 10, fill := 'steelblue') %>%
    add_axis('x', title = '', 
        properties = axis_props(
            grid = list(strokeWidth = 0), 
            ticks = list(strokeWidth = 0), 
            labels = list(angle = 30, align = 'left')
        )
    ) %>%
    add_axis('y', title = 'billions USD', title_offset = 50)
```

## Conclusions

It is interesting to see that tornadoes are so much more deadly than ever other 
type of weather event, and yet it is flooding and hurricanes that do the most 
damage to property and crops. This suggests that water levels that are not high 
enough to cause widespread drowning can still do significant damage to structures 
and fields. Additionally, the speed with which a tornado occurs likely prevents 
prior evacuation, whereas hurricans tend to occur with a few days warning, and 
flooding tends to build up more slowly. This is not the case with flash floods, 
but flash floods tend to occur in low-population areas such as desert 
wash-basins.

The most obvious policy direction suggested by this research is that a very good 
deal, probably most, of the research dollars allocated to storm research should 
be allocated to tornado research, as developing the ability to accurately predict 
these even just a few hours more ahead of time can save more lives than any other 
meteorology-related measure. Additionally, the $2.5 trillion cumulative damage 
done by flooding and storm surges suggests money can be saved by investing in 
levees and discouraging development in low-lying flood basins.
