---
title: "The relationship between transmission and mpg"
author: "Adam Acosta"
output: pdf_document
---

```{r setup, cache=F, echo=F, message=F, warning=F, tidy=F, results='hide'}
library(knitr)
library(xtable)
library(stargazer)
data(mtcars)
attach(mtcars)
opts_chunk$set(message=F, error=F, warning=F, comment=F, dpi=100, tidy=F, 
               echo=F, results="asis", fig.height=5, fig.width=6.5, 
               fig.path="final/figures/")
setwd("/home/adam/datasciencecoursera/regmod")
```

### Executive Summary

We wish to answer the question: Is an automatic or manual transmission better for 
MPG? As well as to quantify the MPG difference between transmission types. We 
use the correlation matrix and scatterplot matrix to cut down the number of 
plausible covariates, then conduct two-way ANOVAs and conclude that a manual 
transmission is better for MPG, but is not statistically significant if we 
control for weight.

### Data Description

The data come from Motor Trend magazine showing miles per gallon and ten other 
variates in 32 cars from the 1974 model year, including engine cylinders, 
displacement ($in^3$), horsepower, rear axle ratio, weight (thousands lbs), 1/4 
mile time, cylinder configuration, transmission, number of gears, and number of 
carburetors. 

### Covariate Selection and Exploratory Analysis

We first select by appealing to theory. Physics tells us that it takes more 
force to move a heavy than light vehicle and a more powerful engine burns more 
fuel. Examine the correlation matrix (Figure 1). 
Note the high correlation between cyl and disp: more cylinders means a larger 
engine. There is a high correlation between these and hp, as large engines tend 
to be more powerful. Note the negative correlation between V/S and cyl, as an 
engine having straight pistons is telling us it is a small engine (no straight 8s 
were produced in 1974). 

We must include wt as a covariate, as this makes the most physical sense. Of the 
measures of power, we use hp, as that among the three is least correlated with 
wt. This also allows us to exclude carb, as the use of multiple carburetors in 
70s muscle cars was a high-performance engine enhancement allowing multiple 
intake manifolds, so is really measuring hp. Finally, note that hp and wt are 
still correlated. So we decouple power from weight by taking the ratio, 
which has a very low correlation, `r round(cor(hp/wt,wt), 2)`. This gives us the 
proposed model. 

$$MPG = \beta_0 + \beta_1 AM + \beta_2 WT + \beta_3 \frac{HP}{WT}$$

### Model Selection and Interpretation

```{r data mutate}
library(dplyr)
newdata <- select(mutate(mtcars,power=round(hp/wt,3)), mpg, wt, am, power)
attach(newdata)
remove(mtcars)
```

First, we model mpg as a function of only wt, the best predictor with the best 
theoretical explanatory power. We compare this simplest model to models that 
include power and transmission using two-way ANOVA (Table 1). The p-value 
in the first comparison suggests it is significant (p = 0.002), but in the second 
comparison is not (p = 0.17). This seems to indicate that power adds explanatory 
value but transmission type does not. Note the rather large F-ratio (12.15) when 
adding power but the rather small F-ratio (1.99) when adding transmission. 

```{r anova1}
fit1 <- lm(mpg ~ wt)
fit2 <- lm(mpg ~ wt + power)
fit3 <- lm(mpg ~ wt + power + am)
table1 <- xtable(anova(fit1, fit2, fit3))
caption(table1) <- "ANOVA wt v. wt + power v. wt + power + am"
print(table1, comment=FALSE, caption.placement="bottom", 
      table.placement="!h")
```

However, if plot mpg against wt by transmission type (Figure 2), each slope 
appears to be different as well as outside the 95% confidence interval of the 
other. This suggests an interaction model, with am as a grouping variable. When 
we try this (Table 2), we get a significant finding (p = 0.009). A q-q 
plot of the residuals ofthe interaction model (Figure 3) shows a reasonably 
normal distribution with no obvious pattern.

```{r anova2}
fit5 <- lm(mpg ~ wt + power + wt*am, data=newdata)
table3 <- xtable(anova(fit2, fit5))
caption(table3) <- "ANOVA wt + power v. wt + power wt*am"
print(table3, comment=FALSE, caption.placement="bottom",
      table.placement="!h")
```

We get the final model comparison shown below (Table 3). 

```{r regression_table}
stargazer(fit1, fit2, fit5, title="Comparison of Regression Models", 
          dep.var.labels=c("Miles per Gallon"),
          covariate.labels=c("Weight (thousands lbs)", 
                             "Horsepower to Weight Ratio",
                             "Transmission (manual = 1)", 
                             "Weight Grouped by Transmission"),
          omit.stat=c("ser","f"), single.row=TRUE, header=FALSE)
```

### Conclusion

We conclude that, although weight is the greatest determinant of fuel efficiency, 
transmission type does have a noticeable effect, amplifying the effect of weight, 
perhaps because a manual transmission is itself lighter than an automatic, so 
weight is not independent of transmission type. The final model we get is:

$$MPG = 34.37 - 3.6 WT - 0.09 \frac{HP}{WT} + 12.4 AM - 3.9 WT \cdot AM$$

Note the rather odd effect (apparent from Figure 2) that the weight penalty for 
having a manual transmissionis greater than for an automatic (-7.7 mpg/1000 lbs 
to -3.6 mpg/1000 lbs), but the manual transmissionadds back on average 12.4 miles 
per gallon by interaction with weight. We offer no explanation, but suggestthat 
future investigation compare like vehicles that are available in both transmissio
n types, to remove allconfounders at the experiment level. It is also possible 
that the relationship between weight and mpg isquadratic, with an increasing 
penalty regardless of transmission type, but the grouping of the data here 
doesnot seem to suggest that, as the relationship is very strongly linear within 
each group.

### Appendix A: Supporting Figures

```{r correlation}
cor_mat <- xtable(round(cor(mtcars),2))
print(cor_mat, comment=FALSE, floating=FALSE)
```

```{r pairs}
#n <- ncol(mtcars)
#op <- par(mfrow=c(n, n), mar=rep(0, 4), oma=c(4.1, 4.1, 5.1, 4.1), 
#          cex.lab=0.85, cex.axis=0.8, cex.main=0.9)
#for(i in 1:n)for(j in 1:n) {
#     xrange = range(mtcars[,j])
#     yrange = range(mtcars[,i])
#     plot.new()
#     plot.window(xlim = xrange, ylim = yrange)
#     if (i == j) {
#          text(mean(xrange), mean(yrange),
#               colnames(mtcars)[j])
#     } else {
#          points(mtcars[,j], mtcars[,i], pch=21, cex=0.5, col="steelblue", 
#                 bg="steelblue")
#          #abline(lm(mtcars[,i] ~ mtcars[,j]), col="tomato3", lwd=0.5)
#     }
#     box(col="lightgray")
#}
#par(op)
```

Figure 1. Correlation matrix of mtcars dataset.

```{r regression}
library(ggthemes)
g <- ggplot(mtcars, aes(wt, mpg, col=factor(am)))
g <- g + geom_point() + geom_smooth(aes(group=factor(am)), method="lm") + 
     theme_few()
g + scale_colour_discrete(name="Transmission", breaks=c(0,1), 
                          labels=c("Automatic","Manual"))
```

Figure 2. Regression lines and confidence intervals of mean for 
$MPG = \beta_0 + \beta_1 WT$ by transmission.

```{r residuals1}
q <- ggplot(mtcars, aes(sample=resid(fit5), colour=factor(am)))
q <- q + stat_qq() + geom_abline(yintercept=0, slope=2, colour="slategray") + 
     theme_few()
q + scale_colour_discrete(name="Transmission", breaks=c(0,1), 
                          labels=c("Automatic","Manual"))
```

Figure 3. q-q plot of model residuals: 
$MPG = 34.37 - 3.6 WT - 0.09 \frac{HP}{WT} + 12.4 AM - 3.9 WT \cdot AM$
